A pipeline to analyse time-course gene expression data

### 2019-08-12 manuscript.tex
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

\begin{verbatim}
##   
   checking for file '/tmp/RtmpwKvOP3/remotes33736f6dc/NelleV-moanin-205e154/DESCRIPTION' ...

v  checking for file '/tmp/RtmpwKvOP3/remotes33736f6dc/NelleV-moanin-205e154/DESCRIPTION'
##

@@

-  building 'moanin_0.0.0.tar.gz' (3.9s)
##

@@

\begin{itemize}
\tightlist
\item
  Do we need installation instructions for moanin in this document, since its
  not on CRAN/bioconductor? Also, how its set up now (doing make file, which
  compiles the documentation, etc), probably prohibits use of, e.g.
  \texttt{install\_github} from \texttt{devtools}. If able to call
  \texttt{install\_github("NelleV/2019timecourse-rnaseq-pipeline",\ subdir="bin/moanin",\ dependencies=TRUE)} would be simpler (would also
  install the dependencies and get around problems in script)
\item
  I got an error in installation script for packages:
\end{itemize}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# Reorder genes on condition, time, and replicate}

\NormalTok{pca_data =}\StringTok{ }\KeywordTok{prcomp}\NormalTok{(}\KeywordTok{t}\NormalTok{(variance_filtered_data), }\DataTypeTok{rank=}\DecValTok{3}\NormalTok{, }\DataTypeTok{center=}\OtherTok{TRUE}\NormalTok{, }\DataTypeTok{scale=}\OtherTok{TRUE}\NormalTok{) }
\NormalTok{percent_var =}\StringTok{ }\KeywordTok{round}\NormalTok{(}\DecValTok{100} \OperatorTok{*}\StringTok{ }\KeywordTok{attr}\NormalTok{(pca_data, }\StringTok{"percentVar"}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

We then plot the two first components, and color each sample by (1) its
condition; (2) its sampling time. We use different markers for each replicate. \textbf{EAP: Unclear what `markers' refers to}
We also plot the second and third components in the second row.

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow=}\KeywordTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{, }\DecValTok{2}\NormalTok{), }\DataTypeTok{mar=}\KeywordTok{c}\NormalTok{(}\FloatTok{2.5}\NormalTok{, }\FloatTok{2.5}\NormalTok{, }\FloatTok{2.5}\NormalTok{, }\FloatTok{2.5}\NormalTok{))}
\KeywordTok{plot}\NormalTok{(}
\NormalTok{    pca_data}\OperatorTok{$}\NormalTok{x[, }\StringTok{"PC2"}\NormalTok{], pca_data}\OperatorTok{$}\NormalTok{x[, }\StringTok{"PC1"}\NormalTok{],}
    \DataTypeTok{col=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Group[meta}\OperatorTok{$}\NormalTok{Group],}
    \DataTypeTok{pch=}\NormalTok{ann_markers}\OperatorTok{$}\NormalTok{Replicate[}\KeywordTok{as.factor}\NormalTok{(meta}\OperatorTok{$}\NormalTok{Replicate)],}
    \DataTypeTok{xlab=}\StringTok{"PC2"}\NormalTok{, }\DataTypeTok{ylab=}\StringTok{"PC1"}\NormalTok{)}
\KeywordTok{plot}\NormalTok{(}
\NormalTok{    pca_data}\OperatorTok{$}\NormalTok{x[, }\StringTok{"PC2"}\NormalTok{], pca_data}\OperatorTok{$}\NormalTok{x[, }\StringTok{"PC1"}\NormalTok{],}
    \DataTypeTok{col=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Timepoint[}\KeywordTok{as.factor}\NormalTok{(meta}\OperatorTok{$}\NormalTok{Timepoint)],}
    \DataTypeTok{pch=}\NormalTok{ann_markers}\OperatorTok{$}\NormalTok{Replicate[}\KeywordTok{as.factor}\NormalTok{(meta}\OperatorTok{$}\NormalTok{Replicate)],}
    \DataTypeTok{xlab=}\StringTok{"PC2"}\NormalTok{, }\DataTypeTok{ylab=}\StringTok{"PC1"}\NormalTok{)}

\KeywordTok{plot}\NormalTok{(}
\NormalTok{    pca_data}\OperatorTok{$}\NormalTok{x[, }\StringTok{"PC2"}\NormalTok{], pca_data}\OperatorTok{$}\NormalTok{x[, }\StringTok{"PC3"}\NormalTok{],}
    \DataTypeTok{col=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Group[meta}\OperatorTok{$}\NormalTok{Group],}
    \DataTypeTok{pch=}\NormalTok{ann_markers}\OperatorTok{$}\NormalTok{Replicate[}\KeywordTok{as.factor}\NormalTok{(meta}\OperatorTok{$}\NormalTok{Replicate)],}
    \DataTypeTok{xlab=}\StringTok{"PC2"}\NormalTok{, }\DataTypeTok{ylab=}\StringTok{"PC3"}\NormalTok{)}
\KeywordTok{plot}\NormalTok{(}
\NormalTok{    pca_data}\OperatorTok{$}\NormalTok{x[, }\StringTok{"PC2"}\NormalTok{], pca_data}\OperatorTok{$}\NormalTok{x[, }\StringTok{"PC3"}\NormalTok{],}
    \DataTypeTok{col=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Timepoint[}\KeywordTok{as.factor}\NormalTok{(meta}\OperatorTok{$}\NormalTok{Timepoint)],}
    \DataTypeTok{pch=}\NormalTok{ann_markers}\OperatorTok{$}\NormalTok{Replicate[}\KeywordTok{as.factor}\NormalTok{(meta}\OperatorTok{$}\NormalTok{Replicate)],}
    \DataTypeTok{xlab=}\StringTok{"PC2"}\NormalTok{, }\DataTypeTok{ylab=}\StringTok{"PC3"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

\NormalTok{variance_filtered_data =}\StringTok{ }\NormalTok{variance_filtered_data[, ord]}
\NormalTok{data_corr =}\StringTok{ }\KeywordTok{cor}\NormalTok{(variance_filtered_data, }\DataTypeTok{method=}\StringTok{"pearson"}\NormalTok{)}
\NormalTok{data_corr_meta =}\StringTok{ }\NormalTok{meta[ord, ]}
\NormalTok{data_corr_meta}\OperatorTok{$}\NormalTok{Timepoint =}\StringTok{ }\KeywordTok{as.factor}\NormalTok{(data_corr_meta}\OperatorTok{$}\NormalTok{Timepoint)}

\KeywordTok{aheatmap}\NormalTok{(}
\NormalTok{    data_corr,}
    \DataTypeTok{Colv=}\OtherTok{NA}\NormalTok{, }\DataTypeTok{Rowv=}\OtherTok{NA}\NormalTok{,}
    \DataTypeTok{annCol=}\NormalTok{data_corr_meta[, }\KeywordTok{c}\NormalTok{(}\StringTok{"Group"}\NormalTok{, }\StringTok{"Timepoint"}\NormalTok{)],}
    \DataTypeTok{annRow=}\NormalTok{data_corr_meta[, }\KeywordTok{c}\NormalTok{(}\StringTok{"Group"}\NormalTok{, }\StringTok{"Timepoint"}\NormalTok{)],}
    \DataTypeTok{annLegend=}\OtherTok{TRUE}\NormalTok{, }
    \DataTypeTok{annColors=}\NormalTok{ann_colors,}
    \DataTypeTok{main=}\StringTok{"Correlation plot"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

\begin{tabular}{l}
\hline
x\\
\hline
M.0-VL.0\\
\hline
M.3-VL.3\\
\hline
M.6-VL.6\\
\hline
M.9-VL.9\\
\hline
M.12-VL.12\\
\hline
M.18-VL.18\\
\hline
M.24-VL.24\\
\hline
M.30-VL.30\\
\hline
M.36-VL.36\\
\hline
M.48-VL.48\\
\hline
M.60-VL.60\\
\hline
M.72-VL.72\\
\hline
M.120-VL.120\\
\hline
M.168-VL.168\\
\hline
\end{tabular}

Then run the differential expression analysis on all of those timepoints
jointly.

@@

\begin{tabular}{l|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r}
\hline
  & M.0-VL.0-pval & M.0-VL.0-qval & M.0-VL.0-lfc & M.3-VL.3-pval & M.3-VL.3-qval & M.3-VL.3-lfc & M.6-VL.6-pval & M.6-VL.6-qval & M.6-VL.6-lfc & M.9-VL.9-pval & M.9-VL.9-qval & M.9-VL.9-lfc & M.12-VL.12-pval & M.12-VL.12-qval & M.12-VL.12-lfc & M.18-VL.18-pval & M.18-VL.18-qval & M.18-VL.18-lfc & M.24-VL.24-pval & M.24-VL.24-qval & M.24-VL.24-lfc & M.30-VL.30-pval & M.30-VL.30-qval & M.30-VL.30-lfc & M.36-VL.36-pval & M.36-VL.36-qval & M.36-VL.36-lfc & M.48-VL.48-pval & M.48-VL.48-qval & M.48-VL.48-lfc & M.60-VL.60-pval & M.60-VL.60-qval & M.60-VL.60-lfc & M.72-VL.72-pval & M.72-VL.72-qval & M.72-VL.72-lfc & M.120-VL.120-pval & M.120-VL.120-qval & M.120-VL.120-lfc & M.168-VL.168-pval & M.168-VL.168-qval & M.168-VL.168-lfc\\
\hline
NM\_009912 & 0.2728476 & 0.4745864 & 2.2018266 & 0.7670600 & 0.8760739 & 1.1576881 & 0.0514865 & 0.1487818 & 2.5397431 & 0.1461489 & 0.3125183 & -0.7613983 & 0.2735411 & 0.4753685 & 1.4786596 & 0.5199686 & 0.7057241 & 1.0838768 & 0.0372513 & 0.1172476 & 1.0099581 & 0.1663710 & 0.3415394 & -2.0803981 & 0.0040198 & 0.0209941 & 1.867318 & 0.0000272 & 0.0003259 & -2.877945 & 0.0000001 & 0.0000025 & -2.200448 & 0.0000000 & 0.0000000 & -2.941491 & 0.0000000 & 0.0000000 & 2.659545 & 0.0000000 & 0.0000000 & -4.846082\\
\hline
NM\_008725 & 0.8508522 & 0.9242869 & 2.0281764 & 0.1390300 & 0.3020872 & 1.4568427 & 0.0034217 & 0.0184668 & 2.3999010 & 0.0011765 & 0.0078218 & -0.5946375 & 0.2218494 & 0.4142375 & 0.8974352 & 0.0490869 & 0.1436339 & 1.0689628 & 0.0001128 & 0.0011050 & -1.1910616 & 0.2358903 & 0.4313519 & 0.8409000 & 0.0025790 & 0.0147499 & 1.463357 & 0.0000005 & 0.0000098 & -3.732241 & 0.0992368 & 0.2386529 & -3.305718 & 0.2451045 & 0.4423855 & -5.832292 & 0.0028580 & 0.0160083 & 5.103924 & 0.3298546 & 0.5356392 & 6.770567\\
\hline
NM\_007473 & 0.0880394 & 0.2192542 & 0.6779618 & 0.9165990 & 0.9592675 & 0.7639279 & 0.6859316 & 0.8250549 & 1.2874572 & 0.4958553 & 0.6863397 & -0.9545717 & 0.9654288 & 0.9838265 & 1.1610592 & 0.0996542 & 0.2393445 & -0.5791692 & 0.3501431 & 0.5558993 & -0.7342619 & 0.9783036 & 0.9900678 & 0.8829109 & 0.9915772 & 0.9963117 & -3.926817 & 0.7895585 & 0.8895260 & -3.706969 & 0.5164039 & 0.7028663 & -2.200317 & 0.0909694 & 0.2244290 & 2.218886 & 0.6982871 & 0.8331540 & -1.433851 & 0.6691637 & 0.8141329 & 7.711609\\
\hline
ENSMUST00000094955 & 0.0081879 & 0.0367991 & 1.1856925 & 0.9353016 & 0.9690575 & -0.6333013 & 0.3745215 & 0.5795488 & 2.5957177 & 0.0050684 & 0.0252354 & -0.6988290 & 0.6729944 & 0.8166022 & 1.0178334 & 0.0948389 & 0.2310919 & 0.8228212 & 0.0253932 & 0.0879330 & 1.1482199 & 0.3657233 & 0.5710470 & -1.3750913 & 0.2915369 & 0.4951562 & -3.968832 & 0.5819450 & 0.7532377 & -1.276993 & 0.6373529 & 0.7923844 & 1.548320 & 0.8614351 & 0.9300767 & -4.288514 & 0.8886550 & 0.9448186 & -4.232771 & 0.1879584 & 0.3707972 & 6.442072\\
\hline
NM\_001042489 & 0.4191775 & 0.6208106 & 1.2123141 & 0.7607212 & 0.8723405 & 0.5992596 & 0.7748772 & 0.8807955 & 1.1476237 & 0.1995820 & 0.3858600 & 2.6613947 & 0.0205303 & 0.0748885 & -0.8188268 & 0.1910652 & 0.3748543 & -0.5443243 & 0.8270177 & 0.9111679 & 2.7899965 & 0.5239967 & 0.7088274 & -1.3479903 & 0.1409795 & 0.3049960 & -6.548630 & 0.4062041 & 0.6090027 & 1.889101 & 0.4968748 & 0.6871546 & -3.392918 & 0.0003701 & 0.0030031 & -6.639293 & 0.0000135 & 0.0001774 & -5.488106 & 0.0256136 & 0.0884953 & -4.976920\\
\hline
NM\_008159 & 0.1231754 & 0.2777536 & 1.1022344 & 0.1169243 & 0.2678638 & 0.7374376 & 0.5781815 & 0.7504472 & 0.9406648 & 0.0059119 & 0.0284812 & -0.8908867 & 0.4381368 & 0.6375400 & 1.2644788 & 0.8998830 & 0.9506910 & -0.5490689 & 0.5634704 & 0.7394176 & -1.4223473 & 0.5015939 & 0.6910094 & -2.3313130 & 0.0000000 & 0.0000000 & 1.321606 & 0.0000000 & 0.0000001 & -2.339338 & 0.0000000 & 0.0000000 & -3.096598 & 0.0000007 & 0.0000123 & -5.350299 & 0.0000125 & 0.0001658 & -6.076895 & 0.0000010 & 0.0000176 & -4.774939\\
\hline
NM\_001013813 & 0.0565287 & 0.1593800 & 0.6299949 & 0.1710643 & 0.3480241 & 0.9746582 & 0.0951123 & 0.2315686 & 3.3644621 & 0.4387357 & 0.6380771 & 0.7308201 & 0.0005373 & 0.0040980 & 1.2229788 & 0.3295522 & 0.5353121 & 1.7341606 & 0.0007915 & 0.0056472 & -1.0267100 & 0.0013222 & 0.0086036 & -0.9327780 & 0.0010519 & 0.0071337 & 1.314872 & 0.1685588 & 0.3445806 & 1.623828 & 0.0554666 & 0.1571666 & -4.828044 & 0.0067413 & 0.0315747 & -5.772946 & 0.0006393 & 0.0047334 & -5.776763 & 0.0004591 & 0.0035968 & -5.017664\\
\hline
AK039774 & 0.0150416 & 0.0590150 & 0.6372239 & 0.0622016 & 0.1707939 & 0.9714819 & 0.0038649 & 0.0203447 & 1.3168792 & 0.9071986 & 0.9543873 & -0.6495320 & 0.0337680 & 0.1089499 & 1.4620498 & 0.1808765 & 0.3613288 & -0.6084973 & 0.4268483 & 0.6276079 & -1.4219238 & 0.8931509 & 0.9471666 & 0.6812377 & 0.3320222 & 0.5378422 & 1.725274 & 0.0000084 & 0.0001177 & 2.756976 & 0.7523631 & 0.8671706 & -4.265131 & 0.0000850 & 0.0008681 & -5.010475 & 0.0000014 & 0.0000238 & -4.623531 & 0.0066328 & 0.0311737 & -4.795976\\
\hline
NM\_013782 & 0.8754233 & 0.9377109 & 0.5940058 & 0.0001107 & 0.0010876 & 1.1803029 & 0.4731855 & 0.6676244 & 1.9274813 & 0.3558245 & 0.5614957 & -0.6697288 & 0.0000009 & 0.0000157 & 1.2377208 & 0.0469473 & 0.1390186 & -1.2969383 & 0.0877739 & 0.2187959 & 0.7832039 & 0.0873966 & 0.2181127 & -0.7705369 & 0.3054160 & 0.5099832 & 1.240504 & 0.0593928 & 0.1651796 & 2.355748 & 0.4210031 & 0.6224586 & -3.879986 & 0.0053367 & 0.0262800 & -5.108367 & 0.0056116 & 0.0273406 & -4.262393 & 0.0042107 & 0.0217896 & -4.617477\\
\hline
NM\_028622 & 0.7667946 & 0.8759265 & 1.6477918 & 0.9457494 & 0.9741115 & -1.8648471 & 0.9903079 & 0.9957216 & 1.1660022 & 0.6332637 & 0.7895226 & 2.1081788 & 0.8592452 & 0.9288844 & 1.2212238 & 0.7658228 & 0.8753935 & -1.1079164 & 0.8331999 & 0.9145679 & -0.8793087 & 0.8359000 & 0.9160123 & -1.2284455 & 0.8894396 & 0.9452334 & 1.160648 & 0.5797431 & 0.7516226 & -3.323953 & 0.5675779 & 0.7424636 & -4.125536 & 0.6699768 & 0.8146554 & -6.126729 & 0.6521300 & 0.8025109 & -4.510424 & 0.8460618 & 0.9215885 & 6.351333\\
\hline
\end{tabular}

We can repeat this for each of the conditions

@@

@@

The distribution of number of genes found differentially expressed by
considering each time-point independantly highlights the challenges of such
approach. Timepoint 6H and 18H of the Kawasaki strain have fewer numbers of
genes found differentially expressed than timepoints 3H and 9H: this is likely
due to biological or technical variances for specific genes at specific
timepoints. \textbf{EAP: We should add more to this}

As a summary, classic differential expression methods are appropriate for
unordered treatments, but fail to use the temporal structure of the data.

@@

As another sanity check, \texttt{moanin} provides a simple utility function to
visualize gene time-course data. Here, we plot the 12 genes with the smallest
p-values.

>>>>

\begin{verbatim}
##   
   checking for file '/tmp/Rtmp730h5L/remotes334541e0f9/NelleV-moanin-867a648/DESCRIPTION' ...

v  checking for file '/tmp/Rtmp730h5L/remotes334541e0f9/NelleV-moanin-867a648/DESCRIPTION'
##

@@

-  building 'moanin_0.0.0.tar.gz' (4.2s)
##

@@

\begin{itemize}
\tightlist
\item
  Why \texttt{dependicies=FALSE}? (same for the README for the repos). Seems easier to just say dependencies=TRUE and avoid the install.R routine.
\item
  I got an error in installation script for packages:
\end{itemize}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{pca_data =}\StringTok{ }\KeywordTok{prcomp}\NormalTok{(}\KeywordTok{t}\NormalTok{(variance_filtered_data), }\DataTypeTok{rank=}\DecValTok{3}\NormalTok{, }\DataTypeTok{center=}\OtherTok{TRUE}\NormalTok{, }\DataTypeTok{scale=}\OtherTok{TRUE}\NormalTok{) }
\NormalTok{percent_var =}\StringTok{ }\KeywordTok{round}\NormalTok{(}\DecValTok{100} \OperatorTok{*}\StringTok{ }\KeywordTok{attr}\NormalTok{(pca_data, }\StringTok{"percentVar"}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

We then plot the two first PC components, and color each sample by (1) its
condition; (2) its sampling time. We use different plotting symbols for each replicate. We also plot the second and third components in the second column.
\textbf{EAP: The colors for timepoint appear to circle around, i.e.~the first and last are very similar colors}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow=}\KeywordTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{, }\DecValTok{2}\NormalTok{))}
\KeywordTok{par}\NormalTok{(}\DataTypeTok{mar=}\KeywordTok{c}\NormalTok{(}\FloatTok{4.5}\NormalTok{, }\FloatTok{4.5}\NormalTok{, }\FloatTok{1.5}\NormalTok{, }\FloatTok{1.5}\NormalTok{))}
\KeywordTok{plot}\NormalTok{(}
\NormalTok{    pca_data}\OperatorTok{$}\NormalTok{x[, }\KeywordTok{c}\NormalTok{(}\StringTok{"PC1"}\NormalTok{,}\StringTok{"PC2"}\NormalTok{)],}
    \DataTypeTok{col=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Group[meta}\OperatorTok{$}\NormalTok{Group],}
    \DataTypeTok{pch=}\NormalTok{ann_markers}\OperatorTok{$}\NormalTok{Replicate[}\KeywordTok{as.factor}\NormalTok{(meta}\OperatorTok{$}\NormalTok{Replicate)], }\DataTypeTok{bty=}\StringTok{"l"}\NormalTok{)}
\KeywordTok{mtext}\NormalTok{(}\DataTypeTok{at=}\DecValTok{275}\NormalTok{,}\DataTypeTok{text=}\StringTok{"Colored by Group"}\NormalTok{, }\DataTypeTok{line=}\FloatTok{0.5}\NormalTok{ ,}\DataTypeTok{font=}\DecValTok{2}\NormalTok{,}\DataTypeTok{adj=}\FloatTok{0.5}\NormalTok{)}
\KeywordTok{legend}\NormalTok{(}\DataTypeTok{x=}\DecValTok{250}\NormalTok{,}\DataTypeTok{y=}\DecValTok{25}\NormalTok{,}\DataTypeTok{legend=}\KeywordTok{names}\NormalTok{(ann_colors}\OperatorTok{$}\NormalTok{Group), }\DataTypeTok{fill=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Group,}\DataTypeTok{bty=}\StringTok{"n"}\NormalTok{,}\DataTypeTok{xpd=}\OtherTok{NA}\NormalTok{,}\DataTypeTok{yjust=}\FloatTok{0.5}\NormalTok{)}
\KeywordTok{par}\NormalTok{(}\DataTypeTok{mar=}\KeywordTok{c}\NormalTok{(}\FloatTok{4.5}\NormalTok{, }\FloatTok{5.5}\NormalTok{, }\FloatTok{1.5}\NormalTok{, }\FloatTok{.1}\NormalTok{))}
\KeywordTok{plot}\NormalTok{(}
\NormalTok{    pca_data}\OperatorTok{$}\NormalTok{x[, }\KeywordTok{c}\NormalTok{(}\StringTok{"PC3"}\NormalTok{,}\StringTok{"PC2"}\NormalTok{)],}
    \DataTypeTok{col=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Group[meta}\OperatorTok{$}\NormalTok{Group],}
    \DataTypeTok{pch=}\NormalTok{ann_markers}\OperatorTok{$}\NormalTok{Replicate[}\KeywordTok{as.factor}\NormalTok{(meta}\OperatorTok{$}\NormalTok{Replicate)],}
    \DataTypeTok{ylab=}\StringTok{""}\NormalTok{, }\DataTypeTok{bty=}\StringTok{"l"}\NormalTok{, }\DataTypeTok{yaxt=}\StringTok{"n"}\NormalTok{)}
\KeywordTok{axis}\NormalTok{(}\DecValTok{2}\NormalTok{, }\DataTypeTok{labels=}\OtherTok{FALSE}\NormalTok{)}
\KeywordTok{par}\NormalTok{(}\DataTypeTok{mar=}\KeywordTok{c}\NormalTok{(}\FloatTok{4.5}\NormalTok{, }\FloatTok{4.5}\NormalTok{, }\FloatTok{1.5}\NormalTok{, }\FloatTok{1.5}\NormalTok{))}
\KeywordTok{plot}\NormalTok{(}
\NormalTok{    pca_data}\OperatorTok{$}\NormalTok{x[, }\KeywordTok{c}\NormalTok{(}\StringTok{"PC1"}\NormalTok{,}\StringTok{"PC2"}\NormalTok{)],}
    \DataTypeTok{col=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Timepoint[}\KeywordTok{as.factor}\NormalTok{(meta}\OperatorTok{$}\NormalTok{Timepoint)],}
    \DataTypeTok{pch=}\NormalTok{ann_markers}\OperatorTok{$}\NormalTok{Replicate[}\KeywordTok{as.factor}\NormalTok{(meta}\OperatorTok{$}\NormalTok{Replicate)], }\DataTypeTok{bty=}\StringTok{"l"}\NormalTok{)}
\KeywordTok{mtext}\NormalTok{(}\DataTypeTok{at=}\DecValTok{275}\NormalTok{,}\DataTypeTok{text=}\StringTok{"Colored by Timepoint"}\NormalTok{,}\DataTypeTok{line=}\FloatTok{0.5}\NormalTok{ ,}\DataTypeTok{font=}\DecValTok{2}\NormalTok{,}\DataTypeTok{adj=}\FloatTok{0.5}\NormalTok{)}
\KeywordTok{legend}\NormalTok{(}\DataTypeTok{x=}\DecValTok{250}\NormalTok{,}\DataTypeTok{y=}\OperatorTok{-}\DecValTok{15}\NormalTok{,}\DataTypeTok{legend=}\KeywordTok{names}\NormalTok{(ann_colors}\OperatorTok{$}\NormalTok{Timepoint), }\DataTypeTok{fill=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Timepoint,}\DataTypeTok{bty=}\StringTok{"n"}\NormalTok{,}\DataTypeTok{xpd=}\OtherTok{NA}\NormalTok{,}\DataTypeTok{yjust=}\FloatTok{0.5}\NormalTok{)}
\KeywordTok{par}\NormalTok{(}\DataTypeTok{mar=}\KeywordTok{c}\NormalTok{(}\FloatTok{4.5}\NormalTok{, }\FloatTok{5.5}\NormalTok{, }\FloatTok{1.5}\NormalTok{, }\FloatTok{.1}\NormalTok{))}
\KeywordTok{plot}\NormalTok{(}
\NormalTok{    pca_data}\OperatorTok{$}\NormalTok{x[, }\KeywordTok{c}\NormalTok{(}\StringTok{"PC3"}\NormalTok{,}\StringTok{"PC2"}\NormalTok{)],}
    \DataTypeTok{col=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Timepoint[}\KeywordTok{as.factor}\NormalTok{(meta}\OperatorTok{$}\NormalTok{Timepoint)],}
    \DataTypeTok{pch=}\NormalTok{ann_markers}\OperatorTok{$}\NormalTok{Replicate[}\KeywordTok{as.factor}\NormalTok{(meta}\OperatorTok{$}\NormalTok{Replicate)],}
    \DataTypeTok{ylab=}\StringTok{""}\NormalTok{, }\DataTypeTok{bty=}\StringTok{"l"}\NormalTok{, }\DataTypeTok{yaxt=}\StringTok{"n"}\NormalTok{)}
\KeywordTok{axis}\NormalTok{(}\DecValTok{2}\NormalTok{, }\DataTypeTok{labels=}\OtherTok{FALSE}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

\NormalTok{variance_filtered_data =}\StringTok{ }\NormalTok{variance_filtered_data[, ord]}
\NormalTok{data_corr =}\StringTok{ }\KeywordTok{cor}\NormalTok{(variance_filtered_data, }\DataTypeTok{method=}\StringTok{"pearson"}\NormalTok{)}
\NormalTok{data_corr_meta =}\StringTok{ }\NormalTok{meta[ord, ]}
\NormalTok{data_corr_meta}\OperatorTok{$}\NormalTok{Timepoint =}\StringTok{ }\KeywordTok{as.factor}\NormalTok{(data_corr_meta}\OperatorTok{$}\NormalTok{Timepoint)}
\CommentTok{# use aheatmap from the NMF package for a heatmap}
\KeywordTok{aheatmap}\NormalTok{(}
\NormalTok{    data_corr,}
    \DataTypeTok{Colv=}\OtherTok{NA}\NormalTok{, }\DataTypeTok{Rowv=}\OtherTok{NA}\NormalTok{,}
    \DataTypeTok{annCol=}\NormalTok{data_corr_meta[, }\KeywordTok{c}\NormalTok{(}\StringTok{"Group"}\NormalTok{, }\StringTok{"Timepoint"}\NormalTok{)],}
    \DataTypeTok{annRow=}\NormalTok{data_corr_meta[, }\KeywordTok{c}\NormalTok{(}\StringTok{"Group"}\NormalTok{, }\StringTok{"Timepoint"}\NormalTok{)],}
    \DataTypeTok{annLegend=}\OtherTok{TRUE}\NormalTok{, }
    \DataTypeTok{annColors=}\NormalTok{ann_colors,}
    \DataTypeTok{main=}\StringTok{"Correlation plot"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

This creates a vector of contrasts to be tested, one for each timepoint:

\begin{tabular}{l}
\hline
Contrasts\\
\hline
M.0-VL.0\\
\hline
M.3-VL.3\\
\hline
M.6-VL.6\\
\hline
M.9-VL.9\\
\hline
M.12-VL.12\\
\hline
M.18-VL.18\\
\hline
M.24-VL.24\\
\hline
M.30-VL.30\\
\hline
M.36-VL.36\\
\hline
M.48-VL.48\\
\hline
M.60-VL.60\\
\hline
M.72-VL.72\\
\hline
M.120-VL.120\\
\hline
M.168-VL.168\\
\hline
\end{tabular}

\textbf{EAP: Not sure if \texttt{kable} is better than just printout?}

Then run the differential expression analysis on all of those timepoints
jointly using the function \texttt{DE\_timepoints}.

@@

\textbf{EAP: Sort of strange to have argument \texttt{spines\_model} for function that doesn't use splines. Perhaps there is a generic name that could be used for the argument}

This results in a table of the results, where each row corresponds to a gene and the columns correspond to the p-value (\texttt{pval}), log-fold change (\texttt{lfc}) and adjusted p-value (\texttt{qval}); the order of the genes in the table is the same as the input \texttt{data}. Here we show the results for the first two timepoints and the first ten genes:

\begin{tabular}{l|r|r|r|r|r|r}
\hline
  & M.0-VL.0-pval & M.0-VL.0-qval & M.0-VL.0-lfc & M.3-VL.3-pval & M.3-VL.3-qval & M.3-VL.3-lfc\\
\hline
NM\_009912 & 0.273 & 0.475 & 2.202 & 0.767 & 0.876 & 1.158\\
\hline
NM\_008725 & 0.851 & 0.924 & 2.028 & 0.139 & 0.302 & 1.457\\
\hline
NM\_007473 & 0.088 & 0.219 & 0.678 & 0.917 & 0.959 & 0.764\\
\hline
ENSMUST00000094955 & 0.008 & 0.037 & 1.186 & 0.935 & 0.969 & -0.633\\
\hline
NM\_001042489 & 0.419 & 0.621 & 1.212 & 0.761 & 0.872 & 0.599\\
\hline
NM\_008159 & 0.123 & 0.278 & 1.102 & 0.117 & 0.268 & 0.737\\
\hline
NM\_001013813 & 0.057 & 0.159 & 0.630 & 0.171 & 0.348 & 0.975\\
\hline
AK039774 & 0.015 & 0.059 & 0.637 & 0.062 & 0.171 & 0.971\\
\hline
NM\_013782 & 0.875 & 0.938 & 0.594 & 0.000 & 0.001 & 1.180\\
\hline
NM\_028622 & 0.767 & 0.876 & 1.648 & 0.946 & 0.974 & -1.865\\
\hline
\end{tabular}

\textbf{EAP: For the output of this, the column names are in format \texttt{M.0-VL.0-pval}, for example. This may be hard to parse algorithmically down the road because you are using the same character for the contrast (\texttt{-}) as the separator for the \texttt{pval}. So for example, if you wanted to split the contrast name from the \texttt{pval} or \texttt{lfc} part, it wouldn't be very easy. Maybe underscore \texttt{\_} would be better separator?}

We will repeat this, comparing each of the treatments to the control (``M'') (code not printed here, as it is a replicate of the above, see accompanying Rmarkdown document).

@@

\textbf{EAP: Should \texttt{timely\_barplot} be a function of \texttt{moanin}?}

@@

Such an analysis can demonstrate some general trends, with clearly more genes
being differentially expressed at later time points, and the Vietnam high-dose
showing perhaps an earlier onset than the low-dose.

However, the distribution of the number of genes found differentially
expressed by considering each time-point independently highlights the
challenges of such approach. We can see that some timepoints have many less
genes found significantly differentially expressed (e.g.~timepoint 6H and 18H
of the Kawasaki strain). While there may be biological differences at those
time points for some genes, it seems unlikely that the large majority of genes
differentially expressed at timepoint 3H stop being differentially expressed
at 6H and then jump back to being differentially expressed at 9H. A more
likely explanation is that there is are some technical or biological artifact
about the samples for 6H that is creating higher variation and thus less
ability to detect significance.

Another difficulty with such an approach is making sense of the general
temporal structure for any particular gene, as different genes will have
different combinations of timepoints DE. For the comparison of the Kawasaki
strain to the control, for example, there are \texttt{length(signifCombos)} genes
found DE in some timepoints, and there are \texttt{length(tabCombos)} different
combinations of timepoints for which they are DE. Some of these make sense,
such as DE in timepoints 48H-168H (\texttt{tabCombos{[}"48,60,72,120,168"{]}} genes), but
many are very fragmentary. For example there are \texttt{tabCombos{[}"48,60,120,168"{]}}
genes which are DE in timepoints 48,60,120,168H, but \emph{not} in the 72H. Many of
these genes are likely to have not made the cutoff for significance in 72H,
but don't show real differences in the overall trend between 48H-168H.

Here are plots of the first 10 such genes, where we can see that some might
show some meaningful changes between 60 and 72H, but others clearly just have
a single replicate that is different or increased variability at 72H.

\begin{center}\includegraphics{reports/manuscript_files/figure-latex/showMissingDEGenes-1} \end{center}

\textbf{EAP: It would be nice to be able to subset in \texttt{plot\_centroids/plot\_genes} to only some of the groups}

As a summary, classic differential expression methods are appropriate for
unordered treatments, but fail to make use of the temporal structure of the data.

@@

As another sanity check, \texttt{moanin} provides a simple utility function to
visualize gene time-course data. Here, we plot the 10 genes with the smallest
p-values.
